---
title: "Peer assessment ML"
author: "Rene int Veld"
date: "Sunday, November 27, 2016"
output: html_document
---

Introduction project

One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it.

In this project, our goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants in an experiment, and then predict the manner in which they did their exercises. This is the "classe" variable in the training set. We will use any of the other variables to predict with. Underneath we will be describing how we built our model, how we used cross validation, and what is the expected out of sample error is, and why we made the choices we have made. 
Furthermore, we will use your prediction model to predict 20 different test cases.

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. 
More information is available from the website here: http://groupware.les.inf.puc-rio.br/har 
(see the section on the Weight Lifting Exercise Dataset).

The HAR (Human Activity Recognition) Dataset contains 5 classes (sitting-down, standing-up, standing, walking, and sitting) collected on 8 hours of activities of 4 healthy subjects. We also established a baseline performance index. 
Read more: http://groupware.les.inf.puc-rio.br/har#ixzz4RhWds5wi

Loading and preprocessing the data

Data: The training data for this project are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
The test data are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv
The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. 
We would like to thank the Pontifical Catholic University of Rio de Janeiro (PUC-Rio), as they have been very generous in allowing their data to be used for this kind of assignment.

We have downloaded the above mentioned 2 files and stored it in local environment for this peer assessment:

```{r}
setwd("~/R/working directory course/predicting activity")

url.train <- 
    "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
url.test <- 
    "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

download.file(url.train, "train.csv")
download.file(url.test, "test.csv")

```


Now we have to read the csv files into R-files for further processing.

```{r}
act.train <- read.csv("train.csv")
act.test <- read.csv("test.csv")

```

We know have a training database of 19662 observations and 160 variables and a testing database of only 20 observations.
We can get an overview of the 6 participants and the number of observations in both sets:

```{r}
table(act.train$user_name)
table(act.test$user_name)

```
We noted that all field names are the same, except for the last field.
In the training set the last field is called Classe, this is the score describing how well the exercise was performed.
In the test set this field is replaced by problem-id, this is the number of the observation of which a prediction has to be made of the score.

```{r}
table(act.train$classe)
table(act.test$problem_id)

```

Now we have to predict the scores of the test set. We noted that many of the 160 columns do not have a valid result (eg. blank or NA or #DIV/0), especially in the test set, so they can be omitted from both sets. These are fields that represent statistical measures:
amplitude
avg
kurtosis
max
min
skewness
stddev
var
All these fields starting with above terms have been deleted in the train and test set to create new sets on which we will make our prediction.



```{r}
x <- names(act.train)
fields.to.select <- subset(x, ! substr(x[],1,3) %in% c("amp","avg","kur","max","min","ske","std","var"))
train <- subset(act.train, select = fields.to.select)

```

This way we only analyse the 60 relevant fields for our prediction.

Some variables are factors, we will transform them:
- new_window is a almost zero variable, we will remove this
- cvtd_timestamp is based on the raw_timestramps, we can remove this
- user_name has to be converted to 6 indicator-values
- and the classe A-E, which we assume to be a Brazilian grade ranging from A to E, can be converted to a numeric between 0 and 10:
Grade	Scale	Grade Description
A	    9 - 10	Excelente (Exellent)	
B	    7 - 9 	Bom (Good)	
C	    5 - 7 	Aceptable (Average)	
D	    3 - 5 	Suficiente (Sufficient)	
E	    0 - 3	Defieciente (Fail)	

```{r}
train$new_window <- NULL
train$cvtd_timestamp <- NULL
train$adelmo <- ifelse(train$user_name=="adelmo",1,0)
train$carlitos <- ifelse(train$user_name=="carlitos",1,0)
train$charles <- ifelse(train$user_name=="charles",1,0)
train$eurico <- ifelse(train$user_name=="eurico",1,0)
train$jeremy <- ifelse(train$user_name=="jeremy",1,0)
# train$pedro <- ifelse(train$user_name=="pedro",1,0) would cause dependency
train$user_name <- NULL
levels(train$classe) <- c(9.5, 8, 6, 4, 1.5)
train$grade <- as.numeric(as.character(train$classe))
train$classe<- NULL
```


We can now choose a prediction method to predict the scores in the test set.
But first have a look at the features.

```{r}
library(caret)
featurePlot(x = train[,c("user_name","raw_timestamp_part_1")] ,y = train$grade, plot = "pairs")
qplot(grade, yaw_arm, color=user_name, data = train)

```


Now try the glm-method

```{r}
# use Generalized Linear Model
modelFit <- train(grade ~ .,method="glm", data=train)
modelFit

# use Generalized Linear Model with preprocessing
modelFit <- train(grade ~ .,method="glm", preProcess="pca", data=train)
modelFit

# use Random Forest Model
# modelFit <- train(grade ~ .,method="rf",data=train) does no work because 5 or fewer unique values (grades)




```
So far the second part of the assessment.
The third part:

Imputing missing values

There are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data. 

We calculated and reported the total number of missing values in the dataset, it appears that if data is missing, then the data of the whole day is missing.
More days in November are missing than days in October.

As it is not clear how to fill the data of the missing days, we take the values of the mean for that 5-minute interval.

We created a new dataset that is equal to the original dataset but with the missing data filled in as explained

We made a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. The values differ from the estimates from the first part of the assignment in the sense that median is now the same as the mean. This can be explained by adding a lot of avarage values. The mean has not changed, which was to be expected. The impact of imputing missing data on the estimates of the total daily number of steps is very limited.

```{r}

act3 <- subset(act1, is.na(steps))
act3$instance = 1
norecords = group_by(act3, date)
norecords = summarise(norecords, 
                       number=sum(instance))
norecords$day <- weekdays(as.Date(norecords$date))
print(norecords)

act4 <- merge(act3, stepintervals, by= "interval")

act4$step <- act4$avg
act2$step <- act2$steps
act4 <- act4[c("interval","date","step")]
act2 <- act2[c("interval","date","step")]
act5 <- rbind(act4,act2)

stepsaday = group_by(act5, date)
stepsaday = summarise(stepsaday, 
                       sumstep = sum(step))
hist(stepsaday$sumstep, xlab="number of steps per day", ylab="number of days", main = "Revised histogram of the steps per day")
mean(stepsaday$sumstep)
median(stepsaday$sumstep)


```
This concludes part 3; I did not have time for part 4 :()